prolfqua - An R-package for Proteomics Differential Expression Analysis.

### 2022-05-30 inst/MyArticle/paper.Rmd
adding changes from Maria

---
title: prolfqua - An R-package for Proteomics Differential Expression Analysis.
author: 
  - Witold E. Wolski $^{\dagger\ddagger}$^[[Functional Genomics Center Zurich (FGCZ)](https://fgcz.ch) - [Swiss Federal Institute of Technology in Zurich](https://ethz.ch) \| [University of Zurich](https://uzh.ch); Winterthurerstrasse 190, CH-8057 Zurich, Switzerland. $^\ddagger$[Swiss Institute of Bioinformatics (SIB)](https://www.sib.swiss/) $^\dagger$Correspondence \texttt{wew@fgcz.ethz.ch}]
  - Paolo Nanni$^{\ast}$
  - Jonas Grossmann$^{\ast\ddagger}$
  - Maria d'Errico$^{\ast\ddagger}$
  - Ralph Schlapbach$^{\ast}$
  - Christian Panse$^{\ast\ddagger}$
output: 
  bookdown::pdf_document2:
      toc: true
      toc_appendix: false
      toc_bib: false
bibliography: bibliography.bib
editor_options: 
  chunk_output_type: console
abstract: |
    Protein mass spectrometry is widely used for quantitative proteomics studies, relative protein quantification, and differential expression analysis of proteins. Nevertheless, there is still a need for a flexible and easy-to-use application programmer interface in R that transparently supports a variety of well principled statistical procedures. The package can model simple experimental designs with a single explanatory variable and complex experiments with multiple factors and hypothesis testing.
    The `r BiocStyle::Githubpkg("fgcz/prolfqua")` R package integrates, using an R6 application programmer interface, essential steps of the mass spectrometry-based differential expression analysis workflow: quality control, data normalization, protein aggregation, statistical modeling, hypothesis testing, and sample size estimation. 
     Furthermore, the package implements benchmark functionality that can help to compare data acquisition, data preprocessing, or data modeling methods using a gold standard dataset. Finally, we show that the implemented methods allow sensitive and specific differential expression analysis. 
     The R package is available through https://github.com/fgcz/prolfqua.
urlcolor: blue
---

@@

Proteins are carriers of function and structure in living cells. Hence, measuring changes in protein abundance is the subject of active research [@vidova2017review]. Bottom-up mass spectrometric methods, where proteins are specifically and reproducible digested into protein fragments - peptides, are employed to identify and quantify proteins in complex samples containing hundreds to thousand of proteins, [@Bubis2017, @da2020philosopher]. The peptides are first separated by their chemical and physical properties using High-Pressure Liquid Chromatography (HPLC). Afterward, they are weighted, identified, and quantified using mass spectrometric techniques, e.g., Electro-Spray-Ionization Mass Spectrometry (ESI-MS). Finally, peptide identification is achieved by fragmenting and matching the measured fragment masses to theoretical masses computed from known peptide sequences[@eng2015deeper, @yu2016pipi, @kong2017msfragger]. For quantification, intact peptide ions [@yu2020fast, @Cox2008MaxQuant] or products of peptide ion fragmentation [@rost2014openswath, @demichev2020dia] are counted and aggregated to obtain precursor abundances. Finally, the identified and quantified peptides are assigned to proteins based on protein sequence information.

Proteomics quantification experiments enable monitoring relative abundances of thousands of proteins in biological samples. Most studies use parallel-group designs, where one or many treatment groups are compared to the control group [@de2022apoe2, @laubscher2021baf]. More recently, more complex experimental designs with an increasing number of samples are studied, e.g., factorial designs and longitudinal studies (time series), sometimes with repeated measurements on the same subject [@tan2022proteomic, @meier2021reduced]. The data can be modeled using linear fixed-, mixed-, or random-effects models [@Bates2015FittingJSS]. Based on these models, tests can be applied to examine whether specific factors and factor interactions are significant, e.g., it can be tested if differences in protein abundance between groups are statistically significant.

An important aspect of mass spectrometric data are missing peptide and protein quantifications. [@rubin1976inference] classified missing data problems into three categories: missing completely at random (MCAR), missing at random (MAR), and missing not at random (MNAR). For instance, in data-dependent acquisition (DDA) MS, only a limited number of MS1 signals are selected for fragmentation and identified. Peptide quantification algorithms transfer identification information between MS1 features in different samples by aligning the data using retention time and mass information, thus reducing the amount of missing data. Nevertheless, highly abundant proteins can suppress the detection of other proteins in a sample, an MAR mechanism. Furthermore, a weak correlation between the number of missing measurements in a condition and the intensities of the observed features can be observed, which is caused by the limit of detection (LOD), an MNAR mechanism [@mcgurk2020use].

Several packages exist to model mass spectrometry protein quantification experiments, e.g., `r BiocStyle::Biocpkg("limma")` [@Ritchie2015], `r BiocStyle::Biocpkg("MSstats")` [@Choi2014], `r BiocStyle::Biocpkg("PECA")` [@Suomi2017bEnhanced],  `r BiocStyle::Biocpkg("msqrob2")` [@Goeminne2016] or `r BiocStyle::Biocpkg("proDA")` [@bioxrvproDA2020], to mention some, all implemented in the R. Each of them has some unique features, for example, `r BiocStyle::Biocpkg("MSstats")` infers the model and generates the model formula from the sample annotation, allowing users with limited statistical knowledge to perform differential expression analysis (DEA). At the same time, `r BiocStyle::Biocpkg("limma")` allows to specify a design matrix using a linear model formula and implements the empirical Bayes variance shrinkage method. The R-package `r BiocStyle::Biocpkg("PECA")`, performs a roll-up of peptide level differences and peptide level $p$-value estimates, obtained from `r BiocStyle::Biocpkg("limma")` or `r BiocStyle::Biocpkg("PECA")`, to protein level estimates. Furthermore, `r BiocStyle::Biocpkg("msqrob2")` implements two models: robust linear models fitted to protein intensities and robust ridge regression fitted to peptide intensities and combines them with empirical Bayes variance shrinkage. The `r BiocStyle::Biocpkg("proDA")` package implements a linear probabilistic dropout model to account for missing data without imputation. `r BiocStyle::Biocpkg("MSstats")` handles missing data by feature filtering and using imputation. Other means of modelling missing observations are the Hurdle model discussed by [@goeminne2020msqrob], while the R package `r BiocStyle::Biocpkg("proDA")` models missingness using probabilistic dropout models [@bioxrvproDA2020].

@@

```{r tableOverview}
xx <- list(
  MSstats = c("pd" = "Y", "rm" = "Y", "mem" = "y"),
  ROPECA = c("pd" = "Y", "rm" = "Y*", fd = "Y", int = "Y", eb = "Y"),
  limma  = c("pd" = "Y", "rm" = "Y*", fd = "Y", int = "Y",eb = "Y"),
  MSqRob2_rlm = c("pd" = "Y", "rm" = "Y*", fd = "Y", int = "Y",eb = "Y", md = "?"),
  proDA = c("pd" = "Y", "rm" = "Y*", fd = "Y", int = "Y",eb = "Y", md = "Y"),
  prolfqua = c("pd" = "Y", "rm" = "Y", fd = "Y", int = "Y", mem = "Y",eb = "Y", md = "Y"))

@@

When developing the R package `r BiocStyle::Githubpkg("fgcz/prolfqua")` we were inspired by the R package `r BiocStyle::CRANpkg("caret")` [@caret2008], which enables to call various machine learning (ML) methods, making selecting the best ML algorithm for your problem easy. We aimed for a similar R package for the analysis of LFQ data. However, after examining the R packages for modeling proteomics quantification data, we observed that model specification, input, and output formats differ widely and wildly. However, what these packages have in common is that they fit linear model either to peptide or protein intensities, determine contrasts among conditions, and afterward, apply empirical Bayes variance shrinkage. Therefore, the revised goal was to provide a modular object oriented (OO) design, with R linear models as a core, where features such as $p$-value aggregation, e.g., `r BiocStyle::Biocpkg("PECA")`, variance shrinkage, or modelling of missing observations can be added.

Furthermore, the functionality of `r BiocStyle::Githubpkg("fgcz/prolfqua")` also includes methods specific to proteomics data. For example, we implemented strategies to estimate protein intensities from peptide intensities: top N, Tukey's median polish, robust linear models [@Grossmann2010, @goeminne2020msqrob]. Furthermore, peptide or protein abundances can be scaled and transformed, using robust scaling, transformed using _quantile_ normalization or `r BiocStyle::Biocpkg("vsa")` to remove systematic differences among samples and heteroscedasticity. In this respect, `r BiocStyle::Githubpkg("fgcz/prolfqua")` can be compared with other R packages such `r BiocStyle::Biocpkg("DEP")` [@DEP2018] or `r BiocStyle::Biocpkg("POMA")` [@POMA2021] which support the entire LFQ analysis pipeline.

We also implemented functionality to streamline the benchmarking of the implemented methods. We use the Ionstar [@shen2018ionstar] dataset to benchmark the modeling methods implemented within `r BiocStyle::Githubpkg("fgcz/prolfqua")` and to compare our results with those of `r BiocStyle::Biocpkg("MSstats")` and `r BiocStyle::Biocpkg("proDA")`. Since group sizes are relatively small, typically with four or five subjects per group, a high power of the tests is a relevant criterion to asses the modelling method. Furthermore, the quantified proteins can be ranked using the estimated fold-change, $t$-statistics, or scaled $p$-value, and then subjected to gene set enrichment (GSEA) or over-representation analysis [@subramanian2005gene] to determine up or down-regulated groups of proteins. However, the model must also provide an unbiased estimate of the false discovery rate (FDR) to manage expectations when selecting protein lists for follow-up experiments.

@@

We store all the data needed for analysis in a single data frame in a tidy table, i.e., every column is a variable, every row is an observation, every cell is a single value[@tidydata2014]. Using an __R6__[@R6cite] configuration object, we specify what variable is in which column, making it easy to integrate new inputs in `r BiocStyle::Githubpkg("fgcz/prolfqua")` if provided as a tidy table. For example, to visualize tidy _Spectronaut_, or _Skyline_ outputs, or data in `r BiocStyle::Biocpkg("MSstats")` format, only a few lines of code are needed to update the `r BiocStyle::Githubpkg("fgcz/prolfqua")` configuration. For popular software like _MaxQuant_ or _MSFragger_, which stores the same variable, e.g., intensity, in multiple columns, one for each sample, we implemented methods that transform the data into tidy tables. Relying on the tidy data table enable us to easily interface with many data manipulation, visualization, and modeling methods implemented in base R[@RCoreTeam2021] and the tidyverse[@tidyverse2019]. We use __R6__ classes to structure the functionality of the package (see Figure \@ref(fig:LFQData) and Figure \@ref(fig:ContrastUML)). __R6__ classes are well supported, e.g., by command-line completion features, see Figure \@ref(fig:tabCompletion), and easy to use (OOP).

@@

\begin{eqnarray} 
\textrm{var} \hat\beta_c &=& c\sigma^2 (X^T X)^{-1} c^T
\end{eqnarray}

@@

The package prolfqua provides function to determine the vector of $parameter$ weights $c$, from a contrast specification string:

```{r eval=FALSE, echo=TRUE}
c("contrast_name" = "factorlevel_1 - factorlevel_2")
```

@@

R linear model and linear mixed effect models allow modeling parallel designs, repeated measurements, factorial designs, and many more. 
Models in `r BiocStyle::Githubpkg("fgcz/prolfqua")` are specified using R’s linear model and mixed model formula interface. Therefore, knowledge of the R regression model infrastructure [@faraway2016extending,@MASS2002] is an advantage when using our package. Furthermore, this glass box approach should make it easy to reimplement an analysis which was performed with `r BiocStyle::Githubpkg("fgcz/prolfqua")` using base R or other programming languages by reading the analysis script. However, acknowledging the R formula interface’s complexity to non-statisticians and the popularity of `r BiocStyle::Biocpkg("MSstats")`, we provide the functionality to suggest a model formula from the sample annotation provided in tabular form similarly to `r BiocStyle::Biocpkg("MSstats")` [@Choi2014].

@@

We focused in our benchmark on the comparison of the statistical modelling methods and fixing the pre-processing steps. However, there are other equally or even more important parameters of a protein quantification pipeline [some paper with workflow evaluation]. One of them is the normalization of the intensities within the samples to remove systematic differences and to make them comparable [@pursiheimo2015optimization]. The method used to infer protein intensities from peptide intensities is an additional important factor [@Grossmann2010]. For instance, the original _proDA_ publication uses MaxLFQ [@cox2014accurate] protein estimates. However, when using MaxLFQ intensities reported by MaxQuant, _proDA_ performs significantly worse ($pAUC_{10}(\textrm{t-statistics}) = 66%$) compared with results obtained when protein abundances are estimated from peptide abundances using Tukey's median polish ($pAUC_{10}(\textrm{t-statistics}) = 73%$). Last but not least, the software [@Cox2008MaxQuant,@yu2020fast], identifying proteins, and generating the quantification values can also significantly contribute to the performance of the entire pipeline, altering the number of identified proteins and the sensitivity and specificity of the differential expression analysis.

@@

\begin{table}[H]
\centering
\begin{tabular}[H]{ll}
\toprule
Abbreviations & Explaination\\
\midrule
AUC    &Area Under the Curve                               \\
CDF    &Cumulative Distribution Function                   \\
ESI-MS &Electro-Spray-Ionization Mass Spectrometry         \\
$FDP$  &False Discovery Proportion                         \\
$FDR$  &False Discovery Rate                               \\
HPLC   &High-Pressure Liquid Chromatography                \\
LC-MS  &Liquid Chromatography followed by Mass Spectrometry\\
LFQ    &Label-Free Quantification                          \\
LOD    &Limit Of Detection                                 \\
MAR    & Missing At Random                                 \\ 
MCAR   & Missing Completely At Random                      \\
MS     &mass spectrometry                                  \\
OO     &Object-Oriented                                    \\
UML     &       Unified Modeling Language                  \\
\bottomrule
\end{tabular}
\end{table}

>>>>

---
title: prolfqua - An R-package for Proteomics Differential Expression Analysis.
author: 
  - Witold E. Wolski $^{\dagger\ddagger}$^[[Functional Genomics Center Zurich (FGCZ)](https://fgcz.ch) - [Swiss Federal Institute of Technology in Zurich](https://ethz.ch) \| [University of Zurich](https://uzh.ch); Winterthurerstrasse 190, CH-8057 Zurich, Switzerland. $^\ddagger$[Swiss Institute of Bioinformatics (SIB)](https://www.sib.swiss/) $^\dagger$Correspondence \texttt{wew@fgcz.ethz.ch}]
  - Paolo Nanni$^{\ast}$
  - Jonas Grossmann$^{\ast\ddagger}$
  - Maria d'Errico$^{\ast\ddagger}$
  - Ralph Schlapbach$^{\ast}$
  - Christian Panse$^{\ast\ddagger}$
output: 
  bookdown::pdf_document2:
      toc: true
      toc_appendix: false
      toc_bib: false
bibliography: bibliography.bib
editor_options: 
  chunk_output_type: console
abstract: |
    Protein mass spectrometry is widely used for quantitative proteomics studies, relative protein quantification, and differential expression analysis of proteins. Nevertheless, there is still a need for a flexible and easy-to-use application programming interface in R that transparently supports a variety of well principled statistical procedures. The prolfqua package can model simple experimental designs with a single explanatory variable and complex experiments with multiple factors and hypothesis testing.
    It is an package that integrates essential steps of the mass spectrometry-based differential expression analysis workflow: quality control, data normalization, protein aggregation, statistical modeling, hypothesis testing, and sample size estimation. 
     Furthermore, the package implements benchmark functionality that can help to compare data acquisition, data preprocessing, or data modeling methods using a gold standard dataset. Finally, we show that the implemented methods allow sensitive and specific differential expression analysis. 
     The R package is available on github https://github.com/fgcz/prolfqua.
urlcolor: blue
---

@@

Proteins are carriers of function and structure in living cells. Hence, measuring changes in protein abundance is the subject of active research [@vidova2017review]. Bottom-up mass spectrometric methods, where proteins are specifically and reproducible digested into protein fragments - peptides, are employed to identify and quantify proteins in complex samples containing hundreds to thousand of proteins, [@Bubis2017; @da2020philosopher]. The peptides are first separated by their chemical and physical properties using High-Pressure Liquid Chromatography (HPLC). Afterward, they are weighted, identified, and quantified using mass spectrometric techniques, e.g., Electro-Spray-Ionization Mass Spectrometry (ESI-MS). Finally, peptide identification is achieved by fragmenting and matching the measured fragment masses to theoretical masses computed from known peptide sequences [@eng2015deeper; @yu2016pipi; @kong2017msfragger]. For quantification, intact peptide ions [@yu2020fast; @Cox2008MaxQuant] or products of peptide ion fragmentation [@rost2014openswath; @demichev2020dia] are counted and aggregated to obtain precursor abundances. Finally, the identified and quantified peptides are assigned to proteins based on protein sequence information.

Proteomics quantification experiments enable monitoring relative abundances of thousands of proteins in biological samples. Most studies use parallel-group designs, where one or many treatment groups are compared to the control group [@de2022apoe2; @laubscher2021baf]. More recently, more complex experimental designs with an increasing number of samples are studied, e.g., factorial designs and longitudinal studies (time series), sometimes with repeated measurements on the same subject [@tan2022proteomic; @meier2021reduced]. The data can be modeled using linear fixed-, mixed-, or random-effects models [@Bates2015FittingJSS]. Based on these models, tests can be applied to examine whether specific factors and factor interactions are significant, e.g., it can be tested if differences in protein abundance between groups are statistically significant.

An important aspect of mass spectrometric data are missing peptide and protein quantifications. [@rubin1976inference] classified missing data problems into three categories: missing completely at random (MCAR), missing at random (MAR), and missing not at random (MNAR). For instance, in data-dependent acquisition (DDA) MS, only a limited number of MS1 signals are selected for fragmentation and identified. Peptide quantification algorithms transfer identification information between MS1 features in different samples by aligning the data using retention time and mass information, thus reducing the amount of missing data. Nevertheless, highly abundant proteins can suppress the detection of other proteins in a sample, an MAR mechanism. Furthermore, a weak correlation between the number of missing measurements in a condition and the abundance of the quantified protein can be observed, which is caused by the limit of detection (LOD), an MNAR mechanism [@mcgurk2020use].

Several packages exist to model mass spectrometry protein quantification experiments, e.g., `r BiocStyle::Biocpkg("limma")` [@Ritchie2015], `r BiocStyle::Biocpkg("MSstats")` [@Choi2014], `r BiocStyle::Biocpkg("PECA")` [@Suomi2017bEnhanced],  `r BiocStyle::Biocpkg("msqrob2")` [@Goeminne2016] or `r BiocStyle::Biocpkg("proDA")` [@bioxrvproDA2020], to mention some, all implemented in the R. Each of them has some unique features, for example, `r BiocStyle::Biocpkg("MSstats")` infers the model and generates the model formula from the sample annotation, allowing users with limited statistical knowledge to perform differential expression analysis (DEA). At the same time, `r BiocStyle::Biocpkg("limma")` allows to specify a design matrix using a linear model formula and implements the empirical Bayes variance shrinkage method. The package `r BiocStyle::Biocpkg("PECA")`, performs a roll-up of peptide level differences and peptide level $p$-value estimates, obtained from `r BiocStyle::Biocpkg("limma")` or `r BiocStyle::Biocpkg("PECA")`, to protein level estimates. Furthermore, `r BiocStyle::Biocpkg("msqrob2")` implements two models: robust linear models fitted to protein intensities and robust ridge regression fitted to peptide intensities and combines them with empirical Bayes variance shrinkage. The `r BiocStyle::Biocpkg("proDA")` package implements a linear probabilistic dropout model to account for missing data without imputation. `r BiocStyle::Biocpkg("MSstats")` handles missing data by feature filtering and using imputation. Other means of modelling missing observations are the Hurdle model discussed by [@goeminne2020msqrob], while the R package `r BiocStyle::Biocpkg("proDA")` models missingness using probabilistic dropout models [@bioxrvproDA2020].

@@

```{r tableOverview}
xx <- list(
  MSstats = c("pd" = "Y", "rm" = "Y", "mem" = "y"),
  ROPECA = c("pd" = "Y", "rm" = "Y*", fd = "Y", int = "Y", eb = "Y"),
  limma  = c("pd" = "Y", "rm" = "Y*", fd = "Y", int = "Y",eb = "Y"),
  MSqRob2_rlm = c("pd" = "Y", "rm" = "Y*", fd = "Y", int = "Y",eb = "Y", md = "?"),
  proDA = c("pd" = "Y", "rm" = "Y*", fd = "Y", int = "Y", eb = "Y", md = "Y"),
  prolfqua = c("pd" = "Y", "rm" = "Y", fd = "Y", int = "Y", mem = "Y",eb = "Y", md = "Y"))

@@

When developing the R package `r BiocStyle::Githubpkg("fgcz/prolfqua")` we were inspired by the R package `r BiocStyle::CRANpkg("caret")` [@caret2008], which enables us to call various machine learning (ML) methods, which makes selecting the best ML algorithm for your problem easy. We aimed for a similar R package for the differential expression analysis of proteomics data. However, after examining the R packages for modeling proteomics quantification data, we observed that model specification, input, and output formats differ widely and wildly. However, these packages have in common: they fit linear models either to peptide or protein intensities, determine contrasts among conditions, and afterward, apply empirical Bayes variance shrinkage. Therefore, the revised goal was to provide a modular object-oriented (OO) design, with R linear models as a core, where we can add features such as $p$-value aggregation, e.g., `r BiocStyle::Biocpkg("PECA")`, variance shrinkage, or modeling of missing observations.

Furthermore, the functionality of `r BiocStyle::Githubpkg("fgcz/prolfqua")` also includes methods specific to proteomics data. For example, we implemented strategies to estimate protein intensities from peptide intensities: top N [@Grossmann2010], Tukey's median polish, robust linear models [@goeminne2020msqrob]. Furthermore, peptide or protein abundances can be scaled and transformed, using robust scaling, transformed using _quantile_ normalization or `r BiocStyle::Biocpkg("vsa")` to remove systematic differences among samples and heteroscedasticity. In this respect, `r BiocStyle::Githubpkg("fgcz/prolfqua")` can be compared with other R packages such as `r BiocStyle::Biocpkg("DEP")` [@DEP2018] or `r BiocStyle::Biocpkg("POMA")` [@POMA2021] which support the entire differential expression analysis pipeline.

We also implemented functionality to streamline the benchmarking of the implemented methods. We use the Ionstar [@shen2018ionstar] dataset to benchmark the modeling methods implemented within `r BiocStyle::Githubpkg("fgcz/prolfqua")` and to compare our results with those of `r BiocStyle::Biocpkg("MSstats")` and `r BiocStyle::Biocpkg("proDA")`. Since group sizes are relatively small, typically with four or five subjects per group, the high power of the tests is a relevant criterion to assess the modeling method. Furthermore, the quantified proteins can be ranked using the estimated fold-change, $t$-statistics, or scaled $p$-value, and then subjected to gene set enrichment (GSEA) or over-representation analysis [@subramanian2005gene] to determine up or down-regulated groups of proteins. Furthermore, the statistical model must provide an unbiased estimate of the false discovery rate (FDR) to manage expectations when selecting protein lists for follow-up experiments. We will use the partial area under the receiver operator curve (ROC) to assess the power of the tests and compare the FDR with the false discovery proportion (FDP).

@@

We store all the data needed for analysis in a single data frame in a tidy table, i.e., every column is a variable, every row is an observation, every cell is a single value [@tidydata2014]. Using an __R6__[@R6cite] configuration object, we specify what variable is in which column, making it easy to integrate new inputs in `r BiocStyle::Githubpkg("fgcz/prolfqua")` if provided as a tidy table. For example, to visualize tidy _Spectronaut_, or _Skyline_ outputs, or data in `r BiocStyle::Biocpkg("MSstats")` format, only a few lines of code are needed to update the `r BiocStyle::Githubpkg("fgcz/prolfqua")` configuration. For popular software like _MaxQuant_ or _MSFragger_, which stores the same variable, e.g., intensity, in multiple columns, one for each sample, we implemented methods that transform the data into tidy tables. Relying on the tidy data table enable us to easily interface with many data manipulation, visualization, and modeling methods implemented in base R [@RCoreTeam2021] and the tidyverse [@tidyverse2019]. We use __R6__ classes to structure the functionality of the package (see Figure \@ref(fig:LFQData) and Figure \@ref(fig:ContrastUML)). __R6__ classes are well supported, e.g., by command-line completion features, see Figure \@ref(fig:tabCompletion), and easy to use (OOP).

@@

\begin{eqnarray} 
\textrm{var} (\hat\beta_c) &=& \sqrt{ c^T \sigma^2 (X^T X)^{-1} c }
\end{eqnarray}

@@

[TODO WEW : provide model formula]

The package `r BiocStyle::Githubpkg("fgcz/prolfqua")` provides function to determine the vector of $parameter$ weights $c$, from a linear models and a contrast specification string.

If the linear model was specified using the following R formula:

```{r}
Intensity ~ factor_1 + factor_2 + factor_1:factor_2
```

then the contrasts ca

```{r eval=FALSE, echo=TRUE}
c("contrast_name" = "factor_1level_1 - factor_1level_2")
```

@@

R linear model and linear mixed effect models allow modeling parallel designs, repeated measurements, factorial designs, and many more. 
Models in `r BiocStyle::Githubpkg("fgcz/prolfqua")` are specified using R’s linear model and mixed model formula interface. Therefore, knowledge of the R regression model infrastructure [@faraway2016extending; @MASS2002] is an advantage when using our package. Furthermore, this glass box approach should make it easy to reimplement an analysis which was performed with `r BiocStyle::Githubpkg("fgcz/prolfqua")` using base R or other programming languages by reading the analysis script. However, acknowledging the R formula interface’s complexity to non-statisticians and the popularity of `r BiocStyle::Biocpkg("MSstats")`, we provide the functionality to suggest a model formula from the sample annotation provided in tabular form similarly to `r BiocStyle::Biocpkg("MSstats")` [@Choi2014].

@@

We focused in our benchmark on the comparison of the statistical modelling methods and fixing the pre-processing steps. However, there are other equally or even more important parameters of a protein quantification pipeline [some paper with workflow evaluation]. One of them is the normalization of the intensities within the samples to remove systematic differences and to make them comparable [@pursiheimo2015optimization]. The method used to infer protein intensities from peptide intensities is an additional important factor [@Grossmann2010]. For instance, the original _proDA_ publication uses MaxLFQ [@cox2014accurate] protein estimates. However, when using MaxLFQ intensities reported by MaxQuant, _proDA_ performs significantly worse ($pAUC_{10}(\textrm{t-statistics}) = 66%$) compared with results obtained when protein abundances are estimated from peptide abundances using Tukey's median polish ($pAUC_{10}(\textrm{t-statistics}) = 73%$). Last but not least, the software [@Cox2008MaxQuant; @yu2020fast], identifying proteins, and generating the quantification values can also significantly contribute to the performance of the entire pipeline, altering the number of identified proteins and the sensitivity and specificity of the differential expression analysis.

@@

\begin{table}
\centering
\begin{tabular}{ll}
\toprule
Abbreviations & Explaination\\
\midrule
AUC    &Area Under the Curve                               \\
CDF    &Cumulative Distribution Function                   \\
ESI-MS &Electro-Spray-Ionization Mass Spectrometry         \\
$FDP$  &False Discovery Proportion                         \\
$FDR$  &False Discovery Rate                               \\
HPLC   &High-Pressure Liquid Chromatography                \\
LC-MS  &Liquid Chromatography followed by Mass Spectrometry\\
LOD    &Limit Of Detection                                 \\
MAR    & Missing At Random                                 \\ 
MCAR   & Missing Completely At Random                      \\
MS     &mass spectrometry                                  \\
OO     &Object-Oriented                                    \\
UML     &       Unified Modeling Language                  \\
\bottomrule
\end{tabular}
\end{table}
